### 2019-07-05 report/chapter02.Rnw
adjustments after correction

% LaTeX file for Chapter 02

<<echo=FALSE>>=
library(knitr)
opts_chunk$set(
    fig.path='figure/ch02_fig',
    self.contained=FALSE,
    cache=TRUE
)
@

<<echo=FALSE>>=
library(knitr)
opts_chunk$set(fig.path='figure/ch02_fig',
               echo=TRUE, message=FALSE,
               fig.width=8, fig.height=3,
               out.width='\\textwidth-3cm',
               message=FALSE, fig.align='center',
               background="gray98", tidy=FALSE, #tidy.opts=list(width.cutoff=60),
               cache=TRUE
)
options(width=74)
@

<<echo = FALSE, warning=FALSE, message=FALSE>>=
rm(list = ls())
PATH_HOME = path.expand("~") # user home
PATH = file.path(PATH_HOME, 'Data/PubBias')
PATH2 = file.path(PATH_HOME, 'PubBias')
FILE = 'cochrane_2018-06-09.csv'
PATH_DATA = file.path(PATH, 'data')
PATH_CODE = file.path(PATH2, 'code')
PATH_RESULTS = file.path(PATH2, 'results')
PATH_FIGURES = file.path(PATH_RESULTS, 'figures')

file_results = "pb.RData"

source(file.path(PATH_CODE, 'PubBias_functions.R'))

file.dat <- "data.RData"
if (file.exists(file.path(PATH_RESULTS, file.dat))) {
	load(file.path(PATH_RESULTS, file.dat))
} else {
	data = pb.readData(path = PATH_DATA, file = FILE)
	tmp = pb.clean(data)
	data = tmp[[1]]
	aliases = tmp[[2]]
	save(data, file =  file.path(PATH_RESULTS, file.dat))
}

load(file.path(PATH_RESULTS, "meta_analyses_summary_bin_cont_surv.RData"))
load(file.path(PATH_RESULTS, "meta_analyses_summary_complete.RData"))
load(file.path(PATH_RESULTS, "meta.bin.RData"))
load(file.path(PATH_RESULTS, "meta.cont.RData"))
load(file.path(PATH_RESULTS, "meta.surv.RData"))
load(file.path(PATH_RESULTS, "data_used_for_analysis.RData"))

###################################################################################################################
###################################################################################################################

#Barbiturate Examples
barbi1 <- arrange(filter(data, file.nr == 21) %>% 
          select(Study = study.name, Comparison = comparison.name, Outcome = outcome.name, 
                 Events = events1, Total = total1, Events_c = events2, Total_c = total2) %>% 
            slice(c(1,3)), Study) 
barbi2 <- arrange(filter(data, file.nr == 21) %>% 
          select(Study = study.name, Comparison = comparison.name,Outcome = outcome.name), Study)
barbi2[barbi2$Study == "P\303\251rez-B\303\241rcena 2008", 1] = "Perez-Barcena 2008"

#Missing values
cont.out <- data %>% filter(outcome.measure.new == "Mean Difference" | outcome.measure.new == "Std. Mean Difference" | 
                                 outcome.measure.new == "Hedges' G")

missing.mean <- which(is.na(cont.out$mean1) | is.na(cont.out$mean2))
missing.effect <- which(is.na(cont.out$effect))
missing.means <- intersect(missing.mean, missing.effect) #Cont. results that have neither complete means nor an effect
wrong.effect <- which(cont.out$mean1 == 0 & cont.out$mean2 == 0 & cont.out$effect == 0) #Cont.results that have only zeros
missing.effects <- length(c(missing.means, wrong.effect))

missing.sd <- which(is.na(cont.out$sd1) | is.na(cont.out$sd2))
missing.se <- which(is.na(cont.out$se))
missing.sds <- intersect(missing.se, missing.sd) #Cont. results that have neither complete means nor an effect
wrong.sds <- which(cont.out$sd1 == 0 & cont.out$sd2 == 0 & cont.out$se == 0) #Cont.results that have only zeros
missing.sds <- length(c(missing.sds, wrong.sds))

missing.ss <- which(is.na(data$total1) | is.na(data$total2))
wrong.ss <- which(data$total1 <= 0 | data$total2 <= 0)
missing.ssize <- length(c(missing.ss, wrong.ss))

missing.year <- sum(is.na(data$study.year)) + length(c(which(data$study.year < 1920), which(data$study.year > 2019)))

missing.table <- rbind("Missing mean values and mean differences" = missing.effects,
                         "Missing standard deviations and standard errors" = missing.sds,
                         "Missing sample sizes" = missing.ssize,
                         "Missing study year" = missing.year)

#Mean and median number of different outcome types of reviews
mean.diff.out <- data %>% group_by(file.nr, comparison.nr) %>% distinct(outcome.name) %>%
  ungroup() %>% group_by(file.nr) %>%
  count() %>% ungroup %>% summarise(mean = mean(n))

median.diff.out <- data %>% group_by(file.nr, comparison.nr) %>% distinct(outcome.name) %>%
  ungroup() %>% group_by(file.nr) %>%
  count() %>% ungroup %>% summarise(mean = median(n))

#Study.year quantiles:
#data.ext2 <- data.ext %>% mutate(study.year = ifelse(study.year > 1920, study.year, NA))
publication.year.range <- c(quantile(data.ext2$study.year, 0.05, na.rm = T), quantile(data.ext2$study.year, 0.25, na.rm = T), 
                            quantile(data.ext2$study.year, 0.5, na.rm = T), quantile(data.ext2$study.year, 0.75, na.rm = T),
                            round(mean(data.ext2$study.year, na.rm = T), 2))

#Outcome.measure frequency table:
outcome.measure.frequencies <- rbind(data %>% group_by(outcome.measure.new) %>% count() %>% 
                                       arrange(desc(n)) %>% ungroup() %>% filter(row_number() < 9),
                                     data %>% group_by(outcome.measure.new) %>% count() %>% arrange(desc(n)) %>% 
                                       ungroup() %>% filter(row_number() > 8) %>% 
                                       summarise(outcome.measure.new = "other", n = sum(n)))

outcome.measure.frequencies <- outcome.measure.frequencies %>% mutate(percentage = round(n/sum(n),3)*100)
names(outcome.measure.frequencies) <- c("Outcome measure", "n", "Percentage")
outcome.measure.frequencies$Percentage <- paste(outcome.measure.frequencies$Percentage, "%", sep = "")

#Total and group size quantiles:
data.ext2 <- data.ext2 %>% ungroup() %>%  mutate(total.n = total1 + total2)
samplesize.range <- c(quantile(data.ext2$total.n, 0.05, na.rm = T), quantile(data.ext2$total.n, 0.25, na.rm = T), 
                            quantile(data.ext2$total.n, 0.5, na.rm = T), quantile(data.ext2$total.n, 0.75, na.rm = T),
                            round(mean(data.ext2$total.n, na.rm = T), 2))
treatment.group.size.range <- c(quantile(data.ext2$total1, 0.05, na.rm = T), quantile(data.ext2$total1, 0.25, na.rm = T), 
                      quantile(data.ext2$total1, 0.5, na.rm = T), quantile(data.ext2$total1, 0.75, na.rm = T),
                      round(mean(data.ext2$total1, na.rm = T), 2))

#Frequencies of results:
comp.freq <- data %>% group_by(file.nr) %>% summarise(n = n())
comp.range <- c(length(which(comp.freq$n < 6)), quantile(comp.freq$n, 0.25, na.rm = T), 
                      quantile(comp.freq$n, 0.5, na.rm = T), quantile(comp.freq$n, 0.75, na.rm = T),
                      round(mean(comp.freq$n, na.rm = T), 2))

#Frequencies of studies:
study.freq <- data %>% group_by(file.nr) %>% distinct(study.name) %>% count()
study.range <- c(length(which(study.freq$n < 3)), quantile(study.freq$n, 0.25, na.rm = T), 
                      quantile(study.freq$n, 0.5, na.rm = T), quantile(study.freq$n, 0.75, na.rm = T),
                      round(mean(study.freq$n, na.rm = T), 2))

#Pooling studies
cum.repr.trials.subg <- data %>% group_by(file.nr, comparison.nr, outcome.nr, subgroup.nr) %>% count %>% group_by(n) %>% count %>%
  filter(n < 15) %>%
  full_join( data %>% group_by(file.nr, comparison.nr, outcome.nr) %>% count %>% group_by(n) %>% count %>%
               filter(n > 14) %>% ungroup %>% summarise(n = 15, nn = sum(nn))) %>% 
  ungroup() %>% arrange(desc(n)) %>% mutate(csum  = cumsum(nn)) %>% arrange(n)
# cum.repr.trials.nosub <- data %>% group_by(file.nr, comparison.nr, outcome.nr) %>% count %>% group_by(n) %>% count %>%
#   filter(n < 15) %>%
#   full_join( data %>% group_by(file.nr, comparison.nr, outcome.nr) %>% count %>% group_by(n) %>% count %>%
#                filter(n > 14) %>% ungroup %>% summarise(n = 15, nn = sum(nn))) %>% 
#   ungroup() %>% arrange(desc(n)) %>% mutate(csum  = cumsum(nn)) %>% arrange(n)

# cum.repr.trials <- cbind(cum.repr.trials.nosub, cum.repr.trials.subg)[,c(1,3,6)]

colnames(cum.repr.trials.subg)  <- c("n","Number of groups", "Cumulative sum of groups")

	data <- data %>% mutate(
		outcome.type = ifelse(outcome.measure.new == "Hazard Ratio", "surv", 
													ifelse(outcome.measure.new == "Odds Ratio" | outcome.measure.new == "Risk Ratio" |
																 	outcome.measure.new == "Peto Odds Ratio" | outcome.measure.new == "Risk Difference", "bin", "cont")),
		outcome.type = ifelse(outcome.measure.new == "Rate Ratio" | outcome.measure.new == "Rate difference", "rate", outcome.type),
		lrr = NA,
		var.lrr = NA,
		smd = NA,
		var.smd = NA,
		smd.pbit = NA,
		var.smd.pbit = NA,
		smd.ordl = NA,
		var.smd.ordl = NA,
		cor.phi = NA,
		var.cor.phi = NA,
		pval.single = NA, 
		cor.pearson = NA,
		var.cor.pearson = NA,
		z = NA,
		var.z = NA,
		events1c = NA,
		events2c = NA,
		sig.single = NA)
	cont.ind <- which(data$outcome.type == "cont" & !is.na(data$outcome.type))

	data[cont.ind, "outcome.type"] <- data[cont.ind, ] %>% 
		mutate(outcome.type = ifelse(outcome.measure.new == "Std. Mean Difference"  | outcome.measure.new == "Mean Difference", 
																 outcome.type, NA)) %>% select(outcome.type)

	cont.ind <- which(data$outcome.type == "cont" & !is.na(data$outcome.type))

	nomean1.ind <- which(data$outcome.type == "cont" & is.na(data$mean1))
	nomean2.ind <- which(data$outcome.type == "cont" & is.na(data$mean2))
	noeffects.ind <- which(data$outcome.type == "cont" & !is.na(data$effect))
	nomeans.ind <- intersect(nomean1.ind, nomean2.ind)
	nomeans.ind <- intersect(noeffects.ind, nomeans.ind) #no means, but a (std.) mean difference is provided.

	data[nomeans.ind,"mean1"] <- data[nomeans.ind,"effect"]
	data[nomeans.ind,"mean2"] <- 0

	data[cont.ind,] <- escalc(data = data[cont.ind,], m1i = mean1, m2i = mean2, 
														sd1i = sd1, sd2i = sd2, n1i = total1, n2i = total2,
														measure = "SMD" , append = T, var.names = c("smd", "var.smd"))

	inpute.given.ind <- which(is.na(data[cont.ind, "smd"]) & data[cont.ind, "outcome.measure.new"] == "Std. Mean Difference")

	#Counts of meta-analyses with only zeros:
all.zero.events.number <- data.ext2 %>% filter(outcome.type == "bin") %>% 
  group_by(meta.id) %>% 
    mutate(n = n()) %>% filter(n >= 10) %>% 
  filter(all(events1 == 0) & all(events2 == 0)) %>% 
  distinct(meta.id) %>% ungroup() %>%  count()

#Counts of meta-analyses with means and sds only zeros:
all.zero.means.number <- data.ext2 %>% filter(outcome.type == "cont") %>% group_by(meta.id) %>% 
    mutate(n = n()) %>% filter(n >= 10) %>% filter(all(mean1 == 0) & all(mean2 == 0)) %>% 
    filter(all(sd1 == 0) | all(sd2 == 0)) %>% distinct(meta.id) %>% ungroup() %>% count()

#Counts of meta-analyses with n larger ten:
m.a.largerten.number <- meta %>% filter(k > 9) %>% ungroup %>% count()
#smaller I2 than 0.5:
m.a.smaller.5.number <- meta %>% filter(k > 9 & I2 < 0.5) %>% ungroup %>% count()
#variance ratio > 4:
m.a.variance4.number <- meta %>% filter(k > 9 & I2 < 0.5) %>% filter((se.max^2)/(se.min^2) > 4) %>% ungroup %>% count()
#At least on sig. estimate:
m.a.one.sig.number <- meta %>% filter(k > 9 & I2 < 0.5) %>% filter((se.max^2)/(se.min^2) > 4) %>% filter(n.sig.single > 0) %>% ungroup %>% count()
#Single patient meta-analyses:
tmp <- data.ext2 %>% filter(outcome.type == "cont") %>% group_by(meta.id) %>% 
	mutate(n = n()) %>% filter(n >= 10) %>% filter(!all(mean1 == 0) & !all(mean2 == 0)) %>% 
	filter(!all(sd1 == 0) | !all(sd2 == 0))

meta.id.vector <- unique(tmp$meta.id)
m.a.singlepatient.number <- length(meta.id.vector[c(which(meta.id.vector > 42716 & meta.id.vector < 42725))] )
#How many z-score meta-analyses are excluded due to total <= 3, and how many m.a. have then n < 10:
tmp <- data.ext2 %>% group_by(meta.id) %>% mutate(n = n()) %>% filter(n > 9) %>%
	filter(outcome.type != "rate" & !is.na(outcome.type))

metac <- meta %>% filter(k > 9) %>% filter(n.sig.single > 0) %>% filter((se.max^2)/(se.min^2) > 4) %>% filter(I2 < 0.5)

meta.id.vector <- metac$meta.id[-which(metac$outcome.type == "surv")]
meta.id.vector <- meta.id.vector[-c(which(meta.id.vector == 157083),
																which(meta.id.vector == 159329),
																# which(meta.id.vector == 12459),
																# which(meta.id.vector == 12465),
																# which(meta.id.vector == 22087),
																which(meta.id.vector == 182298))]

meta.data.zscore <- tmp[which(tmp$meta.id %in% meta.id.vector),]
total.seq3.number <- meta.data.zscore %>% filter(total1 + total2 < 4) %>% select(var.z) %>% ungroup() %>% count() 
m.a.seq3.seq10.number <- meta.data.zscore %>% filter(total1 + total2 > 3) %>% count() %>% filter(nn < 10) %>% ungroup %>% count()

meta.id.vector <- metac$meta.id
meta.data <- data.ext2[which(data.ext2$meta.id %in% meta.id.vector),]

@

\chapter{The Cochrane Dataset}

\section{Cochrane Systematic Reviews}
The Cochrane Group has specialized on systematic reviews in clinical science. Certain knowledge of standards and principles of the Cochrane Group may help to assess the quality and the properties of the dataset. The following information stems from the Cochrane Handbook for Systematic Reviews \citep{cochrane.handbook}. \\
The definition of a systematic review is that it ``attempts to collate all empirical evidence that fits pre-specified eligibility criteria in order to answer a specific research question.'' Thus, the ``key properties of a review are'':

\begin{itemize}
\item``a clearly stated set of objectives with pre-defined eligibility criteria for studies''
\item ``an explicit, reproducible methodology''
\item ``a systematic search that attempts to identify all studies that would meet the eligibility criteria''
\item ``an assessment of the validity of the findings of the included studies, for example through the assessment of risk of bias''
\end{itemize}

At the end of a systematic review, ``a systematic presentation, and synthesis, of the characteristics and findings of the included studies'' is done. \\
53 Cochrane Review Groups prepare and maintain the reviews within specific areas of health care. A group consists of ``researchers, healthcare professionals and people using healthcare services (consumers)''. \\
The groups are supported by Method Groups, Centers and Fields. The Cochrane Method Groups aim to discuss and consult the groups in methodological questions concerning review preparation. The Centers play a main role in training and support of the Groups. The Fields are responsible for broad medical research areas and follow priorities in those areas by advice and control of the groups. \\
The first step in a review is writing a protocol, specifying the research question, the methods to be used in literature search and analysis and the eligibility criteria of the study. Changes in protocols are possible but have to be documented and the protocol is published in advance of the publication of the full review. The choices of methodology as well as the changes should not be made ``on the basis of how they affect the outcome of the research study''. \\
In order to avoid potential conflicts of interests, there is a code of conduct that all entities of the Cochrane Organization have to agree on: conflicts of interest must be disclosed and possibly be forwarded to the Cochrane Center, and participation of review authors in the studies used have to be acknowledged. Additionally, a Steering Group publishes a report of potential conflicts of interests based on information about external funding of Cochrane Groups. \\
In order for keeping the reviews up-to-date, they are revised in a two-year circle with exceptions. In addition to inclusion of new evidence in a field, the revision and maintenance process may as well includes change in analysis methods. This can reflect some advance in clinical science as for example new information about important subgroups, as well as new methods for conducting a Cochrane Review. However, there are no clear guidelines and the Cochrane Groups are free in the rate and extent of up-dating their reviews.

\subsection{Methods for Cochrane Reviews}
A research question defines the following points: ``the types of population (participants), types of interventions (and comparisons), and the types of outcomes that are of interest''. From the research question, usually the eligibility criteria follow. Usually, outcomes are not part of eligibility criteria, except for special cases such as adverse effect reviews. \\
The type of study is an important eligibility criterium. The Cochrane Collaboration focuses ``primarily on randomized controlled trials'', and also, the methods of study identification in literature search are focused on randomized trials. Furthermore, study characteristics such as blinding of study operators with respect to treatment and cluster-randomizing might be additional eligibility criteria which have to be chosen by the review authors. \\

>>>>

% LaTeX file for Chapter 02
<<'preamble04',include=FALSE>>=
library(knitr)
opts_chunk$set(
    fig.path='figure/ch04_fig', 
    self.contained=FALSE,
    cache=FALSE
) 
@

\chapter{Methods} \label{ch:methods}
\section{Introduction and Notation}
The analysis has already been said to be restricted on clinical or health care interventions. The interventions are restricted to comparisons of two treatment arms by some measure of sanity or worsening of health. The difference in this measure between the groups is referred to as the treatment effect. Where it is not particularly mentioned, the term treatment effect refers to any effect measure such as log risk ratio, log hazard ratio, Cohen's $d$ or Hedges $g$, Fisher's z score or Pearsons correlation coefficient. \\
Let us consider a meta-analysis with $n$ study treatment effects ($n > 1$, but typically small). A study is indexed by $i$, and it's treatment effect by  $\theta_i$. The observed treatment effect is $\hat{\theta}_i$. The pooled treatment effect of a meta-analysis will be denoted as $\theta_M$, and consequently, the observed pooled treatment effect as $\hat{\theta}_M$. Furthermore, each treatment effect is typically measured with some standard error $\se_i$ and an estimate of $\se_i$ is denoted as $\hat{\se}_i$. The $\hat{}$ sign thus indicates if it is an estimate.\\
For continuous outcomes, let $m_t$ be the mean of the treatment group, $m_c$ the mean of the control group, and equivalently $\textrm{sd}_t$ and $\textrm{sd}_c$ the corresponding standard deviations. $n_t$ and $n_c$ are the total number of participants in the groups.
In the case of binary outcomes, let $e_t$ be the count of events in the treatment arm $e_c$ the count of events in the control group. The observed counts in a study $i$ are referred to as $e_{t,i}$ and analogously $e_{c,i}$.

\section{Effect Measures and $p$-values}
\subsection{Continuous Outcomes}
For given $(m_t, m_c), (\textrm{sd}_t, \textrm{sd}_c)$ and $(n_t, n_c)$, one can compute mean difference as well as a standardized mean difference (here: Hedges $g$) and  a standard error thereof. Note that the definition of Hedges $g$ and its standard error $\se$ varies among the literature, the following applies for this report:

\begin{align}
\se &= \sqrt{\frac{(n_t - 1)\textrm{sd}_t^2 + (n_c - 1)\textrm{sd}_c^2}{n_t + n_c - 2}} & g &= \frac{m_t - m_c}{\se} \label{eq:hedges.g}
\end{align}

The mean difference $\theta$ and its standard error $\se$ can similarly be obtained by

\begin{align}
\theta &= m_t - m_c & \se &= \sqrt{\textrm{sd}_t^2/n_t + \textrm{sd}_c^2/n_t}
\end{align}

Both estimators take into account that the two groups might have unequal variances. A $p$-value to test the null hypothesis that the mean between group is equal is commonly obtained with the Students $t$ test. The $t$ statistic is obtained, using $s$ and $g$ from \ref{eq:hedges.g}, by

\begin{align}
t &= g/(s\sqrt{(1/n_t)+(1/n_c)}) \nonumber
\end{align}
and the $p$-value can be obtained with the cumulative Student's $t$-distribution $F$ with $n_t + n_c - 2$ degrees of freedom:

\begin{align}
p &= 2(1-F(\abs{t})) \nonumber
\end{align}

The t-test is known to be not very reliable if combined sample size is small ($n < 30$), see for example \citet{t.test}.

\subsection{Binary Outcomes}
Two commonly used effect measures for binary outcome data are risk ratios and odds ratios between treatment and control group. 
The methods presented here can also be found, for example, in \cite[34]{Intro.meta}.
Let $\theta$ be the natural logarithm of the odds ratio. $\theta$ and its variance $\se^2$ can be obtained by

\begin{align}
\theta &= \log(\frac{e_t\cdot(n_c - e_c)}{e_c\cdot(n_t - e_t)}) \nonumber \\
\se^2 &= 1/e_t + 1/(n_t - e_t) + 1/e_c + 1/(n_c - e_c) \nonumber
\end{align}

Plugging in the observed counts will give the corresponding estimates. The logarithm of the risk ratio $\theta$ and its variance $\se^2$ is similarly defined as

\begin{align}
\theta &= \log(\frac{e_t/n_t}{e_c/n_c}) \label{eq:risk.ratio} \\
\se^2&= 1/e_t + 1/n_t + 1/e_c + 1/e_t \label{eq:risk.ratio.variance}
\end{align}

Using likelihood theory, one could show that the estimators are maximum likelihood estimators and that one can use the asymptotic normal distribution of the maximum likelihood estimator to calculate a $p$-value, \eg \cite[98]{held2014}. \\
Thus, with $\Phi$ as the cumulative standard normal distribution, we get
\begin{align}
p &= 2\cdot(1-\Phi(\abs{\hat{\theta}/\hat{\se}}), \nonumber
\end{align}
a $p$-value for the corresponding estimate, which summarizes the evidence against $\hat{\theta}$ being zero (\ie the true risk/odds ratio being 1). \\
Binary and continuous effect measures can be converted into each other, as described in Section \ref{sec:transformation.effectsizes}.

\subsection{Survival Outcomes}
Time-to-event data with censoring has to be analyzed by special means. One frequently used method to take into account right-censoring is the Cox proportional hazards regression model \citep{Cox}. Because the method itself is not applied in this thesis, but only the resulting estimates of the parameters are used, the reader is referred to the extensive literature covering this topic (\eg \citet{Surv}). \\
The so-called hazard ratio estimated by Cox regression is the ratio of the instantaneous risk of experiencing the event between two groups. Because it is a maximum likelihood estimator, one can again use its Wald test statistic to test for equal hazards. Let $\hat{\theta}$ be an estimate of the log hazard ratio and $\hat{\se}$ an estimate of the standard error of it. As before
\begin{align}
p &= 2\cdot(1-\phi(\abs{\hat{\theta}/\hat{\se}(\hat{\theta})}) \nonumber
\end{align}
will give a $p$-value for the evidence against the null hypothesis.

\section{Fixed and Random Effects Meta-Analysis} \label{sec:meta.analysis}
The fixed effects meta-analysis estimator of the pooled treatment effect is a mean of the single treatment effect estimators, weighted by their standard errors \citep{fixed.effects.rosenthal}. Let $w_i = 1/\se_i^2$ be the weights, and $\theta_M$ be the pooled estimator and $s_M^2$ its variance. Then

\begin{align}
\theta_M &= \frac{\sum_{i = 1}^n w_i \theta_i}{\sum_{i = 1}^n w_i} &
s_M^2 &= \frac{1}{\sum_{i = 1}^n w_i} \label{eq:fixed.effects}
\end{align}

This estimator minimizes the variance between the effects. An estimate $\hat{\theta}_M$ can be obtained by plugging in the observed treatment effects and variances $\hat{\theta}_i$ and $\hat{\se}_i^2$. The underlying idea is that we assume $\theta_i \sim N(\theta_M, \se_i^2)$, $\theta_M$ being the true effect, all $\theta_i$ being distributed around an equal mean. \\
The random effects model \citep{whitehead} assumes instead that 
\begin{align}
\theta_i \sim N(\mu_i, \se_i^2) &&
\mu_i \sim N(\theta_M, \tau^2) \label{eq:random.effects} 
\end{align}

Marginally, we have $\theta_i$ being distributed around a common mean $\theta_M$ with additional variance $\tau^2$:

\begin{align}
\theta_i \given \mu_i \sim N(\theta_M, \se_i^2 + \tau^2) \nonumber %\label{eq:random.effects.marginal}
\end{align}

$\tau^2$ is often referred to as a population variance or between-study variance, whereas $\se_i^2$ can be interpreted as sampling error. The pooled treatment effect estimate $\theta_M$ of the random effects model and its variance is obtained by replacing the weights $w_i$ in equation \ref{eq:fixed.effects} with $w_i = 1/(\se_i^2 + \tau^2)$. \\
The model is superior to the fixed effects model whenever the standard errors of the treatment effects alone are unlikely to fully account for the entire variability observed between studies. Note that as $\tau^2$ increases, each $\theta_i$ will eventually get equal weights, irrespective of it's sampling error $\se_i^2$.\\
The estimation of $\tau^2$ has been subject to some debate in the statistical literature. Oftentimes, the method of moment estimator of \citet{tau.estimator} is used.
We use the measure of heterogeneity, $Q$, and divide by $C$ after having subtracted the degrees of freedom $n-1$:

\begin{align}
Q &= \sum_{i = 1}^n w_i(y_i - \theta_M)^2 & C &= \sum_{i = 1}^n w_i - \frac{\sum_{i = 1}^n w_i^2}{\sum_{i = 1}^n w_i}\label{eq:Q.heterogeneity} \\
\tau^2 &= \max\big(0, \frac{Q - (n-1)}{C}\big) \label{eq:Tau.definition}
\end{align}

Again, $w_i, \theta_i ...$ have to be replaced by their estimates in order to get an estimate $\hat{\tau}^2$. \\
The Paule-Mandel estimator \citep{paulemandel} is considered to have most often better properties than the method of moments estimator (\eg \citet{tau.estimator.evaluation}).
Since we defined $w_i = 1/(\se_i^2 + \tau^2)$, it also holds that

\begin{align}
w_i \textrm{Var}(\theta_i) &= 1 & \textrm{Var}(\sqrt{w_i}\theta_i) = 1 \nonumber
\end{align}

For any $w_i$, the variance can be estimated and equated to its expected value:

\begin{align}
\se^2(w_i\theta_i) &= \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} & \frac{\sum_{i = 1}^n w_i(\theta_i - \theta_M)^2}{n-1} = 1 \label{eq:paulemandel}
\end{align}

$\theta_M$ can be estimated using equation \ref{eq:fixed.effects}, the only problem is to estimate $\tau^2$. It can be obtained through an iterative process, using a newly defined function

\begin{align}
F(\tau^2) &= \sum_{i = 1}^n w_i(\theta_i - \theta_M)^2 - (n-1) \nonumber
\end{align}

In view of equation \ref{eq:paulemandel}, $\tau^2$ must be such that $F(\tau^2) = 0$. Then, we start with a arbitrary $\tau^2$ and repeatedly add a term $\tau_0^2$ to update $\tau^2$ until $F(\tau^2 + \tau_0^2)$ is close to zero (using $\tau^2 + \tau_0^2$ for $\hat{w}_i, \hat{\theta}_M$). Using a truncated Taylor series expansion, one can obtain the partial derivative after $\tau^2$, which is a reasonable choice for $\tau_0^2$. \\
%Using $\tau^2 + \tau_0^2$ for $\hat{w}_i, \hat{theta}$, we can update $F(\tau^2)$ and check convergence to zero. \\
The estimation of $\tau^2$ is accompanied by uncertainty. A common procedure is to test if it is there is significant heterogeneity between the studies \cite[109]{Intro.meta}. Compute $Q$, as given in \ref{eq:Q.heterogeneity}, based on one of the estimators of $\tau^2$. It is assumed that $Q$ follows a central Chi-squared distribution with $n -1$ degrees of freedom under the null hypothesis of equally distributed effect sizes. Thus, the expected value of $Q$ is $n-1$, and the excess dispersion is $Q - n + 1$. The $p$-value against the null hypothesis of equally distributed effect sizes is $1 - F(Q)$, using $F$ as the cumulative distribution function of the Chi-squared distribution with d.f. = $n-1$. \\
An advantage of the $\tau^2$ is that it is directly linked to the variability in the data. Additionally, one can use the $I^2$ statistic to see what portion the between study variance has of the overall variance. It is computed as

\begin{align}
I^2 &= \max\Big(0, 1 - \frac{n-1}{Q}\Big) \nonumber%\label{I2.proportion}
\end{align}

%Large values for $\hat{I}^2$ are of great importance when analyzing \eg small study effects. \\
Importantly, all proposed methods above assume normally distributed effect sizes and proper estimates $\hat{\se}$ of the true standard error. This assumptions are not met for very small sample sizes and very few event counts. Alternatively, the Mantel-Haenszel method for risk and odds ratios (see \eg \citet{mantel.haenszel}) could be used in the latter case.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Small Study Effects Tests}
The tests that will be presented on the following pages are a common way to detect publication bias. Importantly, they are however not interpretable directly as evidence for publication bias, but this is left for the discussion chapter. The tests are also often referred to as funnel plot asymmetry tests, because of the popularity of a recent test that has been used frequently to test and adjust for publication bias, which goes under the name of trim-and-fill \citep{trimfill}. However, it is known for some time that the method has disadvantageous properties, therefore, it will not be discussed here, as well as the funnel plot (the radial plot will be used as an alternative).\\
A test for small study effects will test in some ways if the size of the estimated treatment effect of a study depends on some measure of its size. An association between study size and treatment effect size can be interpreted as an artifact of publication bias.

\subsection{Continuous Outcome Tests}
For a continuous outcomes that are normally distributed, the sample mean and variance are independent of each other \cite[120]{meta.w.R}. Thus, the estimated standard errors of treatment effects can be used as a proxy for study size, as they should in principle not be tied to the effect size.

\subsubsection{Begg and Mazumdar: Rank Correlation Test} \label{sec:Begg}
\citet{begg.ties} proposed a rank based test to test the null hypothesis of no correlation between effect size and variance.
A standardized effect size $\theta_i^\star$ can be computed as in \ref{begg.stand.eff}. $\se_i^{2\star}$ is the variance of $\theta_i - \theta_M$ as defined in \ref{begg.stand.var}, $\theta_M$ being the fixed effects pooled treatment effect (\ref{eq:fixed.effects}.

\begin{align}
\theta_i^\star &= (\theta_i - \theta_M)/\se_i^{2\star} \label{begg.stand.eff}  \\
\se_i^{2\star} &= \se_i^2 - 1/\sum_{i = 1}^n\frac{1}{\se_i^2}1 \label{begg.stand.var} 
\end{align}

A rank correlation test based on Kendall's tau is then used. First, the pairs are ordered after their ranks based on $s^{2\star}$. Then, for each $s^{2\star}$ rank, the corresponding ranks based on $\theta^{2\star}$ that are larger are counted and summed up to $u$. The number of ranks based on $\theta^{2\star}$ that are in contrary, smaller, are counted and summed up to $l$. Then the normalized test statistic $Z$ is given as

\begin{align}
Z &= (u - l)/\sqrt{n(n-1)(2n + 5)/18} \nonumber
\end{align}

Thus, large number of concordance between pairs will reflect in large $\hat{u}$ and small $\hat{l}$ and thus lead to a large $\hat{Z}$. The $p$-value is obtained using the standard normal distribution $\Phi$:
\begin{align}
p &= 2\cdot(1-\Phi(\abs{Z})) \nonumber
\end{align}

The changes that have to be made int the case of ties are small and can be found in \cite[410]{begg.ties}.

\subsubsection{Egger's Test: Weighted Linear Regression Test} \label{sec:Egger}
First, the concept of simple linear regression is introduced. In short, the model assumes a dependent variable $y$ to be a linear function of another explanatory variable $x$:

\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon, & \epsilon \sim N(0, \sigma^2) \label{eq:simple.regression}
\end{align}

$\epsilon$ is the residual noise term that becomes necessary when $n$ pairs $(x_i, y_i)$ are given and there is no exact solution. Then it is common to look for the solution that minimizes the squared residuals, the least-squares solution. Formally,

\begin{align}
\operatorname*{argmin}_{\beta_0, \beta_1}\big(\sum_{i = 1}^n y_i - \beta_0 - \beta_1 x_i\big) \label{eq:least.squares}
\end{align}

Let $\mathbf{X}$ be a matrix with the explanatory variables $x$ and $\mathbf{y}$ a corresponding vector for all $y$:
